---
title: "NOAA at a glance"
author: "Chau Pham"
date: "4/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(width = 110)
```

## Synopsis

This project is carried out as part of the course Reproducible Research run by John Hopkins University on Coursera. Some quick cleaning and wrangling are done on the dataset prior to any explorary analyses are taken. From the data, it appears that hurricanes and tornados bring significant damages to the well-being of people and the economic conditions of the US.

-------------------------------------------------------

## Data processing
The NOAA dataset can be downloaded [here][1]. More information is available on the [documentation][2] of the database and the [FAQ][3] page.

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

We first download the bz2 file containing the dataset to our local computer and load into R. **Warning: the dataset is around <span style="color: red;">475 MB</span> after unzipping.**
```{r processing, cache = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "noaa.bz2") ##optional if the dataset has already been downloaded
library(readr)
noaa <- read_csv("noaa.bz2")
```

Having loaded the dataset, let's have a first look at what we are working with.
```{r}
dim(noaa)
```
We have a dataset with 37 variables and more than 900,000 observations. There are a lot of recorded events in the dataset and it is useful to know how many there are, and what types of events occur most frequently.
```{r}
library(dplyr)
noaa$EVTYPE %>% unique() %>% length()
noaa$EVTYPE %>% unique() %>% .[c(1:10, 700:710, 800:810)] 
noaa_freq <- noaa %>% group_by(EVTYPE) %>% tally(sort = TRUE)
noaa_freq
```
There are close to 1,000 types of events recorded. However, at a glance, it appears this gargantuan number signifies at least two problems namely, poor naming convention and typos. To address these two problems, we will need to aggregate different entries of similar types into one category. Another issue we need to address is the existence of "Summary" entries.
```{r category-processing, cache = TRUE}
library(forcats)
noaa$EVTYPE <- toupper(noaa$EVTYPE) 
noaa <- filter(noaa, !grepl("SUMMARY", EVTYPE))  ##Dropping summarising entries

## Categorization
noaa <- noaa %>% mutate(EVENT_CAT = case_when(
    grepl("THUNDE*RE*STORM *W+I*N*D*|^[^NO]*TSTM", EVTYPE) ~ "THUNDERSTORM WIND",
    grepl("HIGH WIND", EVTYPE) ~ "HIGH WIND",
    grepl("^(?=.*WIND)(?!.*THUNDERSTORM)(?!.*TSTM)(?!.*HIGH)", EVTYPE, perl = TRUE) ~ "OTHER WIND",
    grepl("HAIL", EVTYPE) ~ "HAIL",
    grepl("TORNADO", EVTYPE) ~ "TORNADO",
    grepl("FLOOD", EVTYPE) ~ "FLOOD", 
    grepl("SNOW", EVTYPE) ~ "SNOW",
    grepl("RAIN", EVTYPE) ~ "RAIN",
    grepl("LIGHTNING", EVTYPE) ~ "LIGHTNING",
    grepl("HEAT", EVTYPE) ~ "HEAT",
    grepl("COLD", EVTYPE) ~ "COLD", 
    grepl("FUNNEL", EVTYPE) ~ "FUNNEL",
    grepl("FOG", EVTYPE) ~ "FOG",
    grepl("HURRICANE|TYPHOON", EVTYPE) ~ "HURICCANE",
    grepl("BLIZZARD", EVTYPE) ~ "BLIZZARD", 
    grepl("RIP CURRENT", EVTYPE) ~ "RIP CURRENT",
    grepl("WATERSPOUT", EVTYPE) ~ "WATERSPOUT", 
    grepl("TSUNAMI", EVTYPE) ~ "TSUNAMI", 
    grepl("SURF", EVTYPE) ~ "SURF",
    grepl("AVALANCHE", EVTYPE) ~ "AVALANCHE", 
    grepl("DROUGHT", EVTYPE) ~ "DROUGHT",
    grepl("WILDFIRE", EVTYPE) ~ "WILDFIRE",
    grepl("SLEET", EVTYPE) ~ "SLEET",
    TRUE ~ "OTHER"
))

noaa %>% group_by(EVENT_CAT) %>% tally(sort = TRUE)
```
To deal with further categorization and date/time, further cleaning and processing are essential. It is, however, beyond the scope of this RMarkdown project and thus will be left as a future project.

## Results
### 1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?
In order to address the question above, it is crucial we define and identify the variables associated with **population health**. From the dataset, we are mostly interested in variable **FATALITIES** & **INJURIES**.

We can measure the degree of harm by the total number of fatalities and/or injuries done by an event or the average fatalities and/or injuries by that event. These two different measures will likely give two distinct pictures as more frequent events might be less harmful and vice versa. I will make a plot reporting these two measures for events that have at least 100 instances.
```{r harmful-event, fig.align = "center", cache = TRUE, tidy = TRUE}
library(ggplot2)
library(tidyr)
reform <- noaa %>% group_by(EVENT_CAT) %>% filter(n() >= 100) %>% summarise(TOTAL_FATAL = sum(FATALITIES), TOTAL_INJUR = sum(INJURIES), AVG_FATAL = mean(FATALITIES), AVG_INJUR = mean(INJURIES)) %>% pivot_longer(cols = c("TOTAL_FATAL", "TOTAL_INJUR", "AVG_FATAL", "AVG_INJUR"), names_to = "MEASURE") %>% separate(col = MEASURE, into = c("MEASURE", "STATS"), "_") %>% mutate(EVENT_CAT = fct_reorder(EVENT_CAT, value))
reform %>% ggplot(aes(x = EVENT_CAT, fill = STATS)) + geom_bar(aes(y = value), stat = "identity", position = position_dodge()) + coord_flip() + facet_wrap(~ MEASURE, nrow = 1, scales = "free_x") + labs(x = "", y = "", title = "Fatalities and Injuries by event") + scale_fill_discrete(name = "Measure", labels = c("Fatalities", "Injuries")) + theme(plot.title = element_text(hjust = 0.5))
```

If we are interested in the total numbers of injuries and fatalities caused, Tornado is the most harmful, whereas, on average, hurricane causes the most injuries while heat-related events are the most fatal.



### 2. Across the United States, which types of events have the greatest economic consequences?
We calculate the damages in dollars by aggregating estimated damages caused to property and damages caused to crop yields to determine the economic consequences of these events.
```{r economic-consequences, tidy = TRUE, cache = TRUE, fig.align = "center"}
## Leave only entries that have a valid record of damages caused
noaa$PROPDMGEXP <- toupper(noaa$PROPDMGEXP)
noaa$CROPDMGEXP <- toupper(noaa$CROPDMGEXP)
noaa <- noaa %>% filter(PROPDMGEXP %in% c("K", "M", "B", "H", NA), CROPDMGEXP %in% c("K", "M", "B", NA))
noaa <- noaa %>% mutate(MULTIPLIER1 = case_when(
    grepl("K", PROPDMGEXP) ~ 10^3,
    grepl("M", PROPDMGEXP) ~ 10^6,
    grepl("H", PROPDMGEXP) ~ 100,
    grepl("K", PROPDMGEXP) ~ 10^9,
    is.na(PROPDMGEXP) ~ 0
), MULTIPLIER2 = case_when(
    grepl("K", PROPDMGEXP) ~ 10^3,
    grepl("M", PROPDMGEXP) ~ 10^6,
    grepl("K", PROPDMGEXP) ~ 10^9,
    is.na(PROPDMGEXP) ~ 0
)) %>% mutate(PROP_DMG = PROPDMG * MULTIPLIER1, CROP_DMG = CROPDMG * MULTIPLIER2)
reform1 <- noaa %>% group_by(EVENT_CAT) %>% filter(n() >= 100) %>% summarise(TOTAL_PROP = sum(PROP_DMG, na.rm = TRUE),  TOTAL_CROP = sum(CROP_DMG, na.rm = TRUE), AVG_PROP = mean(PROP_DMG, na.rm = TRUE), AVG_CROP = mean(CROP_DMG, na.rm = TRUE)) %>% pivot_longer(cols = 2:5, names_to = "MEASURE") %>% separate(col = MEASURE, into = c("MEASURE", "TYPE"), "_") %>% mutate(EVENT_CAT = fct_reorder(EVENT_CAT, value)) 
reform1 %>% ggplot(aes(x = EVENT_CAT, fill = TYPE)) + geom_bar(aes(y = value), stat = "identity") + coord_flip() + facet_wrap(~ MEASURE, nrow = 1, scales = "free_x") + labs(title = "Economic impact of events", subtitle = "Damages measured in US Dollar", x = "", y = "") + scale_fill_discrete(name = "Types of damage", labels = c("Crop", "Property")) + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```
In aggregate, flood and tornado have done exorbitant amount of damages to the US economy, totalling close to \$150 billions and slightly more than \$75 billions. The former disproportionately affects crop yields and harvests while the latter affects property more. On average, however, hurricane is by far the most destructive with each occurrence causing close to $60 millions in damages, mostly in property. 
