---
title: "NOAA Analysis"
author: "Chau Pham"
date: "4/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

#Analysis of the NOAA Storm Database
##Synopsis
... Placeholder for now

##Data processing
The NOAA dataset can be downloaded [here][1]. More information is available on the [documentation][2] of the database and the [FAQ][3] page.
[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

We first download the bz2 file containing the dataset to our local computer and load into R. **Warning: the dataset is around <span style="color: red;">475 MB<\span> after unzipping.**
```{r processing, cache = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "noaa.bz2") ##optional if the dataset has already been downloaded
library(readr)
noaa <- read_csv("noaa.bz2")
```

Having loaded the dataset, let's have a first look at the dataset.
```{r}
summary(noaa)
dim(noaa)
head(noaa)
```
